---
title: Analysis of variance
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = F, warning = T, error = T, 
                      fig.width = 12, fig.height = 6)
# Library
library('magrittr'); library('tidyr')
# Seed
set.seed(0)
```

---

class: center middle inverse

# One-way Anova

---

# Analysis of variance (Anova)

Used to test differences in **means** of **two or more groups**.

$H_{0}$: Means of all groups are equal.   
$H_{1}$: Mean of at least one group is different from others.

Hence, Anova does not indicate which group/groups is/are different.

Can also be used as a model ( $Y=x+\varepsilon$ ) to determine the effect of different groups of factor variable $x$ on dependent variable $Y$.

In R: use `aov()` and then get the summary of the result with `summary()`.

---

# Assumptions of Anova

Observations are independent, so can not apply to paired data.

Dependent variable is continuous and follows a normal distribution.

The variance of dependent variable is similar in groups.

---

We can specify relationship as a model:

$$Y_{ij}=\mu+\varepsilon_{ij},$$ 

where 

- $Y$ is the value for observation $i$ from group $j$, 
- $\mu$ the overall mean, and 
- $\varepsilon_{ij}$ the difference between overall mean $\mu$ and value for observation $i$ from factor $j$.

---

We can also separate factor $\alpha$ from other variables:

$$Y_{ij}=\mu+\alpha_{j}+\varepsilon_{ij},$$

- $Y$ is the value for observation $i$ from group $j$, 
- $\mu$ the overall mean, 
- $\alpha_{i}$ the difference between mean of group $j$ and overall mean $\mu$, and 
- $\varepsilon_{ij}$ the difference between group mean and value for observation $i$ from factor $j$.

---

# Example data

Multi-environment trial of sorghum at 3 locations across 5 years. We'll assume that the data isbase on random samples.

```{r}
sorg <- agridat::adugna.sorghum
sorg <- sorg[sorg$year == 2004, ]
head(sorg)
```

---

# Example problem

Is mean yield different among locations? Does location have an effect on yield?

```{r, echo = F}
boxplot(yield ~ loc, sorg)
points(tapply(sorg$yield, sorg$loc, mean), pch = '_', cex = 3, col = 'tomato')
```

---

$$Y_{ij}=\mu+\alpha_{j}+\varepsilon_{ij},$$


```{r, echo = F}
mkPlot <- function() {
  plot(yield ~ jitter(as.numeric(loc)), sorg, xaxt = 'n', xlim = c(.5,3.5), 
       xlab = "Location", ylab = "Yield")
  axis(1, at = 1:3, labels = levels(sorg$loc))
  points(tapply(sorg$yield, sorg$loc, mean), pch = '_', cex = 3, col = 'tomato')
  abline(h = mean(sorg$yield), col = 'tomato', lwd = 2, lty = 'dashed')
  legend('topright', 
         c("Group means", "Overall mean"), 
         col = c('tomato'), 
         lty = c('solid', 'dashed'))
}
mkPlot()
```

---

```{r}
mkPlot()
points(c(NA, sorg$yield[1], NA), bg =  'tomato', pch = 21, cex = 2)
```

---

```{r}
sorg$yield[1]
mu <- mean(sorg$yield)
alpha <- mean(sorg$yield[sorg$loc == 'Melkassa']) - mean(sorg$yield)
epsilon <- sorg$yield[1] - mean(sorg$yield[sorg$loc == 'Melkassa'])
mu + alpha + epsilon
```

---

There are two sources of variation: 

- random variation within a group, i.e. $\varepsilon_{ij}$
- variation between groups, i.e. $\alpha_{j}$

The idea behind Anova is comparing these. If the variation between groups is higher that variation within groups by a certain degree, we can say that there's an effect.

---

We use *sums of squares* to quantify these and total variability:

- sum of squares of errors (SSE)  
    $SSE=\sum_{j=1}^{k}\sum_{i=1}^{n} (y_{ij}-\overline{y_{j}})^{2}$, $df = n-k$
- sum of squares between groups (SSA)  
    $SSA=\sum_{j=1}^{k}\sum_{i=1}^{n_{j}} (\overline{y_{j}}-\overline{y})^{2}$, $df = k - 1$
- total sum of squares (SST=SSE+SSA)  
    $SST=\sum_{j=1}^{k}\sum_{i=1}^{n} (y_{ij}-\overline{y})^{2}$, $df = n-1$

---

We use *mean squares* to calculate the value of F-statistic.

- mean squares for SSE (MSE)
    $MSE = SSE / (n - k)$
- mean squares for SSA (MSA)
    $MSA = SSA / (k - 1)$
    
F-statistic is the ratio between these: $F = MSA / MSE$. 

If MSA is substantially higher than MSE, the value of F-statistic is higher and its probability on an F-distribution lower.

---

# Example

Is mean yield different among locations? Does location have an effect on yield?

```{r}
(av <- summary(aov(yield ~ loc, sorg)))
```

---

``` {r, fig.height = 5}
curve(df(x, df1 = av[[1]]['Df'][1, ], df2 = av[[1]]['Df'][2, ]), 1, 40)
abline(v = av[[1]]['F value'][1,1], col = 'tomato')
```

The probability of acquiring an F-statistic this high or higher by chance is very small.

---

class: center middle inverse

# Two-way Anova

---

# Two-way Anova

We can add multiple factors to Anova model. This allows us to consider the iteraction between multiple factors.

There are three types of two-way Anova methods, which differ in how interactions are handled. R's `aov` uses type I. 

$$Y_{ij}=\mu+\alpha_{i}+\beta_{k}+\varepsilon_{ij}$$

---

# Example

Does mean yield differ among locations and trials?

```{r, echo = F}
boxplot(yield ~ loc + trial, sorg)
abline(h = mean(sorg$yield), col = 'tomato', lwd = 2, lty = 'dashed')
points(c(tapply(sorg$yield, list(sorg$loc, sorg$trial), mean)), pch = '_', cex = 3, col = 'tomato')
```

---

``` {r}
summary(aov(yield ~ loc, sorg))
summary(aov(yield ~ loc + trial, sorg))
```

---

class: inverse