---
title: Correlation
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = F, warning = T, error = T)
# Libraries
library('DiagrammeR')
# Set seed
set.seed(0)
```

---

class: center middle inverse

# Correlation

---

# What is correlation?

Relationship between two **continuous** variables.

Indicates the direction of relationship on a scale -1 ... 1 where 0 means lack of relationship.

Does **not** indicate causality.

---

# Causality

.pull-left[
$X$ is correlated to $Y$ , so we might think that $X$ causes $Y$ .
```{r, echo = F}
grViz("digraph {rankdir=LR; node [shape = circle] X -> Y}")
```
]

--

.pull-right[
Actually both might be caused by $Z$ , which leads to the correlation between $X$ and $Y$. 
```{r, echo = F}
grViz("digraph {rankdir=LR; node [shape = circle] X -> Y [dir=none]; Z -> X; Z -> Y}")
```
]

---

# Example

People with high education have high salaries. Does education increase salary?

--

Probably both are influenced by personal characteristics, aptitude, motivation, talent etc.

```{r, echo = F}
grViz("digraph {rankdir=LR; node [shape = circle] 
      Education -> Salary [dir=none]; 
      Personality -> Education; 
      Personality -> Salary}")
```

---

# Pearson’s correlation coefficient

Evaluates **linear** relationship: when x increases by 1 unit, y also increases by 1 unit.

$$r = \frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})}{\sum_{i=1}^{n}(x_i-\bar{x})^2 \sum_{i=1}^{n}(y_i-\bar{y})^2}$$

We compare differences from mean for values of each group.

---

Statistical significance can also be estimated by finding the probability of t-statistic on t-distribution.

$$t=r\sqrt{n-2}/\sqrt{1-r^{2}}$$

---

# Spearman rank-order correlation coefficient

Evaluation is based on the ranking of values. Evaluates a **monotonic** relationship: when x increases, y also increases.

$$r = 1 - \frac{6\sum (x_{i}-y_{i})^{2}}{n(n^{2}-1)}$$

We compare rankings of values from each group.

---

# Pearson and Spearman coefficients

Monotonically increasing linear relationship

```{r, echo = F}
x <- rnorm(100, 100)
y <- x + rnorm(100, 0, .5)
plot(x, y)
legend('topright', 
       legend = c(paste("Pearson's r =", cor(x, y, method = 'pearson') %>% round(2)), 
                  paste("Spearman's r =", cor(x, y, method = 'spearman') %>% round(2))))
```

---

Montonically increasing non-linear relationship

```{r, echo = F}
x <- runif(100, 1, 100)
y <- 100 * x^.1 + rnorm(100)
plot(x, y)
legend('topright', 
       legend = c(paste("Pearson's r =", cor(x, y, method = 'pearson') %>% round(2)), 
                  paste("Spearman's r =", cor(x, y, method = 'spearman') %>% round(2))))
```

---

# How to use in R?

Example data: Nebraska farm income in 2007 by county.

```{r}
farmin <- agridat::nebraska.farmincome
head(farmin, 3) 
```

- `county`, county
- `crop`, crop income, thousand dollars
- `animal`, livestock and poultry income, thousand dollars
- `area`, area of each county, square miles

---

Is income from crops related to income from livestock and poultry?

```{r}
par(pty = 's') # Force the plot area to be square
plot(farmin$crop, farmin$animal)
```

---

To get a coefficient of correlation, use function `cor()`. If data has missing values, add `use = 'pairwise.complete.obs'` as an argument to only use observations where both values are not missing. 

```{r}
cor(farmin$crop, farmin$animal, use = 'pairwise.complete.obs')
```

--

By default `cor()` computes Pearson's correlation coefficient. Use `method = 'spearman'` to calculate Spearman's version.

```{r}
cor(farmin$crop, farmin$animal, use = 'pairwise.complete.obs', method = 'spearman')
```

---

When you pass multiple data frame columns to the function, `cor()` returns a correlation table. 

```{r}
corTab <- cor(farmin[, c('animal', 'crop', 'area')], use = 'pairwise.complete.obs')
corTab
```

---

You can get a visual overview of correlations when you pass a correlation table to `heatmap()` function.

```{r, fig.height = 5}
heatmap(corTab)
```

---

class: inverse