---
title: Statistical hypothesis testing. T-test.
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = F, warning = T, error = T, 
                      fig.width = 12, fig.height = 6)
# Seed
set.seed(0)
```

---

class: center middle inverse

# Inferential statistics

---

## Population and sample

When we have data on the entire population we can simply describe it using *descriptive statistics*.

But often we only have a part of the *population*, i.e. a *sample*.

Since we only have data on sample, we can't be sure if same properties hold for population.

Hence, we need to make *predictions* or *inferences* about the *population*.

We can make inferences with some limited certainty, i.e. confidence.

To do that, we need to use ideas from *probability theory*.

---

# Random sample

Sample must be taken from the population *randomly*.

For this, every value in the population must have an equal chance of being in sample.

E.g. when examining crops, you should not take samples from only one part of the field.

---

# Why statistical hypothesis testing?

Are these samples different? Are these mean values for two samples different?

```{r, echo = F}
makePlot <- function() {
  plot(c(10, 35), c(0, .25), type = 'n', yaxt = 'n', 
       xlab = 'Values', ylab = 'Probability')
  abline(v = c(20, 25), col = 'tomato')
}
makePlot()
```

---

If these were the observed distributions, we could say that means are different.

```{r, echo = F}
makePlot()
lines(density(rnorm(100, 20, 2)))
lines(density(rnorm(100, 25, 2)))
```

---

However, here we can't be so sure.

```{r, echo = F}
makePlot()
lines(density(rnorm(100, 20, 6)))
lines(density(rnorm(100, 25, 6)))
```

---

What we really wish to know: do these samples come from the same population (blue)?

```{r, echo = F}
makePlot()
lines(density(rnorm(100, 20, 4)))
lines(density(rnorm(100, 25, 6)))
lines(density(rnorm(100, 22.5, 4)), col = 'skyblue', lwd = 2)
```

---

## Normal distribution

Commonly used continuous probability distribution due to *central limit theorem*. Many methods assume random variables to be normally distributed.

Symmetric and bell-shaped. Characterized by parameters mean ( $\mu$ ) and variance ( $\sigma^2$ ). ( ${\mathcal{N}}(\mu,\sigma^{2})$ ). For standard normal distribution mean is 0 and variance 1 ( ${\mathcal{N}}(0,1)$ .

---

```{r}
par(mfrow = c(1,3))
curve(dnorm(x, 0, 1), -3,3);curve(dnorm(x, 0, .2), -3,3);curve(dnorm(x, 0, 2), -3,3)
```

---

Some test statistics are assumed to have normal distribution. We can use this to see how probable a value of a test statistic is, e.g. is it at least 95%.

```{r, echo = F}
curve(dnorm(x, 0, 1), -3,3)
abline(v = c(-qnorm(.975), qnorm(.975)), col = 'tomato')
```

---

## How is a test statistic used

The value of a test statstic indicates how extreme the differences are.

E.g. t-statistic used in t-test.

$$t = \frac{\overline{x} - \mu_{0}}{s/\sqrt{n}}$$

---

### Example

Does this wheat come from a population where mean yield is 600 g/plot?

```{r}
wheat <- agridat::wiebe.wheat.uniformity
mean(wheat$yield)
(tScore <- (mean(wheat$yield) - 600) / (sd(wheat$yield) / sqrt(nrow(wheat))))
```

---

What is the probability of getting this extreme t-statistic if this wheat came from a population with mean yield of 600 or more?

.pull-left[
``` {r}
curve(dt(x, nrow(wheat) - 1), from = -5, to = 5)
abline(v = tScore, col = 'tomato')
```
]

.pull-right[
``` {r}
(pVal <- pt(tScore, nrow(wheat) - 1))
```
]

The p-value, hence the probability is very small.

---

## P-value

The probability that a sample value (e.g. difference in means) is equal to or more extreme than true value if null hypothesis is true.

We estimate this probability through a test statistic.

---

## Hypotheses

We only test null hypothesis ( $H_{0}$ ). 

When we can not confirm $H_{0}$, we accept the alternative hypothesis ( $H_{1}$ ).

For testing we set a treshold for p-value **beforehand**, conventionally 0.05.

If p-value is =>0.05, we accept $H_{0}$. If p-value is <0.05, we reject $H_{0}$ and accept $H_{1}$.

**P-value is NOT the probability of $H_{1}$ being true!**

---

# Multiple comparisons problem

Every time you find a statistically significant result, there is a possibility that you got this extreme data by chance. In this case, you committed a type 1 error.

Therefore, if you do a lot of tests, the possibility that at least one of the statistically significant results is actually not true. 

**The number of tests you do should be limited.**

---

class: center middle inverse

# Types of tests

---

## Number of groups

Between how many groups do we estimate differences? 

There are different methods to compare...
- one group and a constant, 
- two groups, 
- three or more groups.

---

### One group and constant

```{r, echo = F}
curve(dnorm(x, 20,3), 10, 30, xlab = "Value", ylab = "Probability")
abline(v = c(20, 25), col = 'tomato')
```

---

### Two groups

```{r, echo = F}
plot(c(10, 40), c(0, .15), xlab = "Value", ylab = "Probability", type = 'n')
lines(density(rnorm(100, 20, 4)))
lines(density(rnorm(100, 30, 4)))
abline(v = c(20, 30), col = 'tomato')
```

---

### Three or more groups

```{r, echo = F}
plot(c(10, 40), c(0, .15), xlab = "Value", ylab = "Probability", type = 'n')
lines(density(rnorm(100, 20, 4)))
lines(density(rnorm(100, 25, 4)))
lines(density(rnorm(100, 30, 4)))
abline(v = c(20, 25, 30), col = 'tomato')
```

---

## Parametric and non-parametric tests

Parametric tests assume that data comes from a popluation with a known distribution (e.g. normal). We can then use **a fixed set of parameters** (e.g. mean and variance) to test differences.

Non-parametric tests do not assume any distribution. We can then only use **rankings of values** to test differences.

Rule of thumb: when your values are normally distributed, use a parametric test, otherwise a non-parametric test.

---

## Paired or unpaired values

Do the pairwise values represent measurements of the same subject? If yes, then values are paired.

Paired data is uncommon, e.g. measurement of the same subject at different times, or two measurements of different phenomena of the same subject.

---

Example of paired data: Milk yield of cows in different periods (`P1`, `P2` and `P3`).

```{r, echo = F}
milk <- agridat::patterson.switchback
milk <- milk[, c('cow', 'period', 'y')]
milk <- tidyr::spread(milk, period, y)
head(milk, 12)
```

---

Example of paired data: Weight of vines (`vines`) and peas (`peas`) in same plot.

```{r, echo = F}
peas <- agridat::nonnecke.peas.uniformity
head(peas, 12)
```

---

## One-tailed or two-tailed tests

What is the relationship that we are testing? 

Do we need to know the probability that parameter values for groups are equal or that one is larger?

---

$H_{1}$: Means of groups are not equal.

```{r, echo = F}
curve(dnorm(x), -4, 4)
polygon(c(-4, seq(-4, -1.96, .01), -1.96), 
        c(0, dnorm(seq(-4, -1.96, .01)), 0), 
        col = 'tomato')
polygon(c(1.96, seq(1.96, 4, .01), 4), 
        c(0, dnorm(seq(1.96, 4, .01)), 0), 
        col = 'tomato')
```

---

With one-tailed tests the test statistic does not have to be as extreme. With 95% confindence intervals, instead of 2.5%, we're now using 5% of extreme values. E.g. a t-statistic value of 1.7 is now significant. 

.pull-left[
$H_{1}$: Mean of one group is smaller.
```{r, echo = F}
curve(dnorm(x), -4, 4)
polygon(c(-4, seq(-4, -1.645, .01), -1.645), 
        c(0, dnorm(seq(-4, -1.645, .01)), 0), 
        col = 'tomato')
```
]

.pull-right[
$H_{1}$: Mean of one group is greater.
```{r, echo = F}
curve(dnorm(x), -4, 4)
polygon(c(1.645, seq(1.645, 4, .01), 4), 
        c(0, dnorm(seq(1.645, 4, .01)), 0), 
        col = 'tomato')
```
]

---

class: center middle inverse

# T-test in R

---

## T-test

Used to compare the mean of one group to a constant or means of two groups.

Assumes that (means of the) values are *independent and identically distributed* (i.i.d.) and follow an (approximately) normal distribution.

Based on the calculation of t-statistic and its location on t-distribution.

---

```{r}
apples <- agridat::archbold.apple
head(apples)
```

--

We assume that we only have a sample of apple trees and we would like to make inferences about the entire population of apple trees.

---

```{r}
plot(range(apples$yield, na.rm = T), c(0,.01), type = 'n')
lines(density(apples$yield, na.rm = T), lwd = 2)
lines(density(apples$yield[apples$gen == 'Golden'], na.rm = T), col = 'orange')
lines(density(apples$yield[apples$gen == 'Redspur'], na.rm = T), col = 'red')
```

---

## Compare group mean to a constant

$H_{0}:$ Mean of group is equal to a constant.  
$H_{1}:$ Mean of group is different from a constant.

In R: `t.test(x, mu)`.

---

Is the mean of Golden apples 160 g/tree?

```{r}
t.test(apples$yield[apples$gen == 'Golden'], mu = 160)
```

--

P-value is less than 0.05. We thus reject $H_{0}$ and accept the $H_{1}$ that the mean yield of Golden apples is not 160 kg/tree.

---

## Compare means of two groups

$H_{0}:$ Means of groups are equal.  
$H_{1}:$ Means of groups are different.

In R: `t.test(y ~ x, data)` or `t.test(x, y, data)`.

---

Is there a difference between the means yields of Golden and Redspur apple trees?

```{r}
t.test(yield ~ gen, apples)
```

--

P-value is greater than 0.05. We thus accept $H_{0}$ that there is no difference between the mean yields of Golden and Redspur.

---

# Two-sided vs one-sided t-test

Tests on previous slides were two-sided. To do a one-sided test, we need to specify the side using argument `alternative` (hypothesis).

Test if mean of `x` is greater than `mu`: `t.test(x, mu, alternative = 'greater')`.

Test if mean of 1st group is less than mean of 2nd group. `mu`: `t.test(y ~ x, data, alternative = 'less')`.

---

Is the yield of Golden apple trees more than 130 kg/tree?

``` {r}
t.test(apples$yield[apples$gen == 'Golden'], mu = 130, alternative = 'greater')
```

--

P-value is less than 0.05. We thus reject $H_{0}$ and accept the $H_{1}$ that the mean yield of Golden apples is higher than 130 kg/tree.

---

Is the yield of Golden apple trees higher than that of Redspur apple trees?

```{r}
t.test(yield ~ gen, apples, alternative = 'greater')
```

--

P-value is greater than 0.05. We thus accept $H_{0}$ that mean yield for Golden is less than or equal to that of Redspur.

---

# Unpaired vs paired t-test

Tests on previous slides were unpaired (default for t-test in R). For a paired test, we need to add argument `paired = TRUE`.

`t.test(y ~ x, paired = FALSE)`

---

Suppose that our data includes only a sample. Are corn yields of US states different when we compare two years?

``` {r}
corn <- agridat::nass.corn
head(corn, 3)
mean(corn$yield[corn$year == 2000])
mean(corn$yield[corn$year == 2010])
```

---

``` {r}
t.test(yield ~ year, corn[corn$year %in% c(2000, 2010), ], paired = T)
```

--

P-value is less than 0.05. We thus reject $H_{0}$ and accept the $H_{1}$ that the corn yields for two years are different.

---

class: inverse