---
title: Non-parametric tests
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = F, warning = T, error = T, 
                      fig.width = 12, fig.height = 6)
# Library
library('magrittr'); library('tidyr')
# Seed
set.seed(0)
```

## Example data

Infection in wheat by different strains of Fusarium. 

```{r}
fus <- agridat::snijders.fusarium
head(fus, 3)
```

- `gen`, wheat genotype
- `strain`, fusarium strain
- `year`, year
- `y`, percent infected

---

```{r, fig.height = 5}
par(mfrow = c(1,4))
for (i in unique(fus$strain)) hist(fus$y[fus$strain == i], 30, xlim = c(0,70), main = i)
fus$f39 <- fus$strain == 'F39'
```

---

## Can we use parameters for testing?

When our (population) values does not follow normal distribution, we can't use tests based on parameters such as mean and variance. 

```{r, echo = F}
hist(fus$y, 30, freq = F, xlab = "Percent infected")
lines(density(fus$y), col = 'skyblue', lwd = 2)
abline(v = mean(fus$y), col = 'tomato', lwd = 2)
legend('topright', "Mean of percent infected", col = 'tomato', lty = 'solid')
```

---

class: center middle inverse

# Mann–Whitney U test

---

## Mann–Whitney U test

Also known as Wilcoxon rank sum test.

Groups are independent, i.e. data is unpaired.

$H_{0}$: Distributions of both samples are the same.  
$H_{1}$: Distributions of samples are different.

In R: `wilcox.test(y ~ x)`. Specify the alternative hypothesis with `alternative`.

---

Formal definition of $H_{0}$: A randomly selected value from one sample is equally likely to be less than or greater than a randomly selected value from a second sample.

``` {r, echo = F}
hist(fus$y, 30, freq = F)
lines(density(fus$y[fus$f39]), col = 'tomato', lwd = 2)
lines(density(fus$y[!fus$f39]), col = 'skyblue', lwd = 2)
legend('topright', c("F39", "Others"), col = c('tomato', 'skyblue'), lty = 'solid')
```

---

## Calculation

```{r}
fusMWU <- rbind(head(fus[fus$f39, c('strain', 'y')]), head(fus[!fus$f39, c('strain', 'y')]))
fusMWU$rank <- rank(fusMWU$y)
fusMWU
```

---

``` {r}
aggregate(rank ~ strain, fusMWU, function(x) sum(x) - (length(x)*(length(x)+1))/2)
```

Smaller of these values is the W-statistic. This is expected to be approximately normally distributed.

---

```{r}
wilcox.test(y ~ strain, fusMWU, paired = F)
```

Value rankings of two groups are different, so they have different distributions.

---

class: center middle inverse

# Wilcoxon signed rank test

---

## Wilcoxon signed rank test

Used for repeated measurements, i.e. data is paired.

$H_{0}$: Distributions of both samples are the same.  
$H_{1}$: Distributions of samples are different.

In R: `wilcox.test(y ~ x, paired = TRUE)`.

---

## Calculation

```{r}
fusWSR <- spread(fus[fus$year %in% 1987:1988, ], year, y, sep = '.')
fusWSR <- fusWSR[1:12, ]
fusWSR$diff <- fusWSR$year.1987 - fusWSR$year.1988
fusWSR$abs.diff <- abs(fusWSR$diff)
fusWSR$rank <- rank(fusWSR$abs.diff)
fusWSR$sign <- sign(fusWSR$diff)
```

---

``` {r} 
fusWSR[-3]
```

---

```{r}
aggregate(rank ~ sign, fusWSR, sum)
```

Smaller of these values is the W-statistic. This is expected to be approximately normally distributed.

---

```{r}
wilcox.test(fusWSR$year.1987, fusWSR$year.1988, paired = T)
```

Differences summed by signs are very different, distributions are not the same.

---

class: center middle inverse

# Kolmogorov-Smirnov test

---

## Kolmogorov-Smirnov test

Compares the overall shape of two distributions, not mean or other parameters. Uses cumulative distribution function to do this.

$H_{0}$: Distributions of both samples are the same.  
$H_{1}$: Distributions of samples are different.

In R: `ks.test(x, y)`.

---

## Cumulative distribution function (CDF)

```{r, echo = F}
par(mfrow = 1:2)
plot(density(fus$y), main = "Probability density function (PDF)")
plot(ecdf(fus$y), main = "Cumulative distribution function (CDF)")
```

---

## Calculation

Test statistic D is the **maximum aboslute difference** between two cumulative distribution functions.

```{r, echo = F}
plot(c(0, 100), 0:1, type = 'n', xlab = "Percent infected")
fus$y[fus$f39] %>% ecdf %>% lines(col = 'tomato')
fus$y[!fus$f39] %>% ecdf %>% lines(col = 'skyblue')
legend('topright', c("F39", "Others"), fill = c('tomato', 'skyblue'))
```

P-value is determined by the extremity of the D-statistic on Kolmogorov distribution.

---

Is the infection rate between strain F39 and other strains different?

```{r}
ks.test(fus$y[fus$f39], fus$y[!fus$f39])
```

--

The maximum aboslute difference between the CDFs is relatively high, so the p-value is very low.

---

class: inverse

---

exclude: true

# Kruskal-Wallis Rank Sum Test

Unpaired, non-parametric

$H_{0}$: Location parameters of the distribution of x are the same in each group.
$H_{1}$: Location parameters of the distribution of x differ in at least one.

`kruskal.test(x ~ y)`

---

exclude: true

# Friedman Rank Sum Test

Repeated measure, within subjects

$H_{0}$: apart from an effect of blocks, the location parameter of y is the same in each of the groups.
$H_{1}$:

`friedman.test()`