---
title: Non-parametric tests
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

## Wilcoxon signed rank test

Non-parmetric

Paired 

`wilcox.test(paired = T)`

---

## Wilcoxon rank sum test

Non-parmetric

Also, Mann–Whitney U test.

Unpaired

$H_{0}$:
$H_{1}$:

`wilcox.test(paired = F)`

---

exclude: true

## Kruskal-Wallis Rank Sum Test

Unpaired, non-parametric

$H_{0}$: Location parameters of the distribution of x are the same in each group.
$H_{1}$: Location parameters of the distribution of x differ in at least one.

`kruskal.test(x ~ y)`

---

exclude: true

## Friedman Rank Sum Test

Repeated measure, within subjects

$H_{0}$: apart from an effect of blocks, the location parameter of y is the same in each of the groups.
$H_{1}$:

`friedman.test()`

---

class: inverse


# Chi-squared test

```{r}
webw <- agridat::beall.webworms
webw$spray <- webw$spray == 'Y'
webw$worms <- webw$y > 1
conTab <- table(webw$spray, webw$worms, dnn = c('Sprayed', 'Webworms present'))
conTab
```


Chi-squared test statistic:

$$\chi^2 = \sum_{i = 1} \frac{n_{obs, i} - n_{expected, i}}{n_{expected, i}}$$

Basically, sum of all differences between observed and expected values.

---

```{r}
conTab
chisq.test(conTab)
```

P-value of `r chisq.test(conTab)$p.value` is below treshold of 0.05, so there's a statistically significant relationship.

---

``` {r, fig.height = 5}
curve(dchisq(x, df = chisq.test(conTab)$parameter), from = 0, to = 100)
abline(v = qchisq(.95, df = chisq.test(conTab)$parameter), col = 'red') # Critical value
abline(v = chisq.test(conTab)$statistic, col = 'blue') # Our value
```

It's very unlikely to get a test statistic with value `r chisq.test(conTab)$statistic` or greater if there was no relationship.