---
title: Ordinary least squares regression
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = F, warning = T, error = T)
# Set seed
set.seed(0)
```

---

class: center middle inverse

# Ordinary least squares regression

---

# Example problem

Weight gain calves in a feedlot, given three different diets.

```{r}
calves <- agridat::urquhart.feedlot
head(calves, 3)
```

- `animal`, animal ID
- `herd`, herd ID
- `diet`, diet: Low, Medium, High
- `weight1`, initial weight
- `weight2`, slaughter weight

---

# Example problem

.pull-left[
Initial and slaughter weight of calves.
```{r, echo = F}
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight")
```
]

--

.pull-right[
Does 1st weight have an effect on 2nd weight?

How much of 2nd weight depends on 1st weight?

How much does 2nd weight increase when 1st weight increases?

What could be the 2nd weight when 1st weight is e.g. 700?

Can we quantify this relationship?
]

---

We can model this using regression!

```{r, echo = F}
wtMod <- lm(weight2 ~ weight1, calves)
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight")
lines(calves$weight1, 
      predict(wtMod, data.frame(weight1 = calves$weight1)), 
      col = 'tomato', lwd = 2)
```

---

# Ordinarly least squares

Also called OLS.

Simplest regression model.

Works only for relationships that are linear (in parameters).

---

# Model specification

Simple linear regression model has one predictor.

$$Y = \alpha + \beta x + \varepsilon,$$

where

- $Y$ is dependent or explained variable, **response** or regressand, 
- $\alpha$ is the intercept or constant, 
- $\beta$ is a coefficient, 
- $x$ is independent or explanatory variable, **predictor** or regressor, and
- $\varepsilon$ is the model error.

---

# Calculation

Minimization of (squared) residuals.

Model parameters are calculated (unlike maximum likelihood estimation based on iterations). 

To estimate the model $Y = \alpha + \beta x + \varepsilon$ :

$\hat{\beta} = \frac{\sum{x_{i} y_{i}} - \frac{1}{n} \sum{x_{i}}\sum{y_{i}}}{\sum{x_{i}^{2}} - \frac{1}{n} (\sum{x_{i}})^{2}} = \frac{Cov[x,y]} {Var[x]}$

$\hat{\alpha} = \overline{y} - \beta \overline{x}$

---

For a model $y = \beta x + \varepsilon$ we can simply use matrix algebra on values of $x$ and $y$:

$$\hat{\beta} = (X^{T} X)^{-1} X^{T} Y$$

In our example problem $X$ would be the vector of 1st weights.

```{r}
calves$weight1
```

And $Y$ would be the vector of 2nd weights.

```{r}
calves$weight2
```

---

# Estimating an OLS model in R

.pull-left[
Function `lm()` (linear model).

The model is defined as a formula.

$Y = \alpha + \beta x + \varepsilon$

In R this is expressed as:  
`y ~ x`  
or  
`y ~ x + z`
]

.pull-right[
```{r}
wtMod <- lm(weight2 ~ weight1, calves)
summary(wtMod)
```
]

---

class: center middle inverse

# Some elements of OLS models

---

# Intercept

.pull-left[
In the formula the $\alpha$.Sometimes called constant. The value of $y$ where the value of $x$ is zero ( $y|x=0$ ).

In our example 2nd weight is `r round(wtMod$coef[[1]])` if 1st weight is 0. Intercept not need to be theoretically valid but it sometimes is.
]

.pull-right[
```{r, echo = F}
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight", 
     xlim = c(0, max(calves$weight1)), ylim = c(0, max(calves$weight2)))
lines(c(-100, calves$weight1), 
      predict(wtMod, data.frame(weight1 = c(-100, calves$weight1))), 
      col = 'tomato', lwd = 2)
abline(v = 0)
points(0, wtMod$coef[1], col = 'steelblue', cex = 4, lwd = 5)
```
]

---

# Coefficient(s)

.pull-left[
In formula the $\beta$. Indicates how much y increases when $x$ increases. In our example every kg of 1st weight increases 2nd weight by `r round(wtMod$coef[[2]], 3)` kg (in the plot *100).

Coefficients are only relevant when their difference from 0 is statistically significant.
]

.pull-right[
```{r, echo = F}
par(pty = 's')
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight")
lines(calves$weight1, 
      predict(wtMod, data.frame(weight1 = calves$weight1)), 
      col = 'tomato', lwd = 2)
predWt <- function(x) predict(wtMod, list(weight1 = x))
lines(c(700, 800), c(predWt(700), predWt(700)))
lines(c(800, 800), c(predWt(700), predWt(800)), col = 'steelblue', lwd = 5)
```
]

---

We can not be sure if the coefficients are actually significant, especially when estimation is done on a sample. 

It is thus necessary to test whether or not coefficients are different from 0. This is done by calculating t-statistic from coefficient and standard error.

---

# Residuals

.pull-left[
In formula the $\varepsilon$. Residuals are the difference of response between observed and fitted values

We use residuals to evaluate how well model fits data. If residuals are large, the model is not very good.
]

.pull-right[
```{r, echo = F}
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight")
lines(calves$weight1, 
      predict(wtMod, data.frame(weight1 = calves$weight1)), 
      col = 'tomato', lwd = 2)
for (i in 1:nrow(calves)) {
  lines(c(calves$weight1[i], calves$weight1[i]), 
        c(calves$weight2[i], predWt(calves$weight1[i])), 
        col = 'steelblue', lwd = 2)
}
```
]

---

# Fitted values

.pull-left[
These are the values of $y$ calculated using the model for every $x$ in the data. 

In other words, predictions.
]

.pull-right[
```{r, echo = F}
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight")
lines(calves$weight1, 
      predict(wtMod, data.frame(weight1 = calves$weight1)), 
      col = 'tomato', lwd = 2)
points(calves$weight1, predWt(calves$weight1), pch = 20, col = 'steelblue', lwd = 5)
```
]

---

# The $R^2$ 

A *goodness of fit* measure: measures how well model fits data. Indicates the **part of variation in response variable that is explained by the model**:

$$R^{2} = \frac{ESS}{TSS} = 1 - \frac{RSS}{TSS}$$

- **e**xplained sum of squares, $ESS$
  $\sum_{i = 1}^{n} (\hat{y}_{i} - \overline{y})^2$
- **r**esidual sum of squares, $RSS$
  $\sum_{i = 1}^{n} (y_{i} - \hat{y}_{i})^2$
- **t**otal sum of squares, $TSS$
  $\sum_{i = 1}^{n} (y_{i} - \overline{y}_{i})^2$

---

How much better is model at explaining the variance of $y$ compared to just the mean?

```{r, echo = F, fig.width = 16}
par(mfrow = 1:2)
# Model
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight", 
     main = "Unexplained by model")
lines(calves$weight1, 
      predict(wtMod, data.frame(weight1 = calves$weight1)), 
      col = 'tomato', lwd = 2)
for (i in 1:nrow(calves)) {
  lines(c(calves$weight1[i], calves$weight1[i]), 
        c(calves$weight2[i], predWt(calves$weight1[i])), 
        col = 'steelblue', lwd = 2)
}
# Mean
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight", 
     main = "Unexplained by mean")
abline(h = mean(calves$weight2), col = 'tomato', lwd = 2)
for (i in 1:nrow(calves)) {
  lines(c(calves$weight1[i], calves$weight1[i]), 
        c(calves$weight2[i], mean(calves$weight2)), 
        col = 'steelblue', lwd = 2)
}
```

---

# The adjusted- $R^2$ 

The more variables we add, the more the model explains.

To penalize a model for the number of predictors ( $K$ ) while considering the number of observations ( $N$ ), the adjusted $R^{2}$ can also be used, particularly for model comparison:

$$\overline{R^{2}} = 1 - \frac{RSS/(N-K)}{TSS/(N-K)}$$

---

# These are in the output of  `summary.lm()`

```{r}
summary(wtMod)
```

---

How are the parameters related to the formula?

$$weight2_{i} = `r round(wtMod$coef[[1]])` + `r round(wtMod$coef[[2]], 3)` * weight1_{i} + \varepsilon_{i}$$

```{r}
calves$weight2[1]
wtMod$coef[[1]] + wtMod$coef[[2]] * calves$weight1[[1]] + wtMod$resid[[1]]
```

---

class: center middle inverse

# Assumptions and diagnostics

---

# Residuals are normally distributed

.pull-left[
```{r}
hist(wtMod$residuals, 30)
```
]

.pull-right[
``` {r}
plot(wtMod, 2)
```
]

---

# Residuals have equal variance

```{r}
plot(wtMod, 1)
```

Here the we have equal variance, e.g. homoscedasticity.

---

Example of heteroscedasticity. Here higher values have higher variance than lower.

```{r, echo = F, fig.width = 16}
par(mfrow = 1:2)
x <- runif(100, 1, 100)
y <- x + x * rnorm(100, 0, .2)
plot(x, y)
abline(lm(y ~ x), col = 'red')
plot(lm(y ~ x), 1)
```

---

# There is no multicollinearity

May be problem when model has multiple predictors.

It should not be possible to linearly predict any of the predictors from others predictors. Otherwise the coefficients will not be reliable.

This can be detected with variance inflation factor by using $Rpredictor._{2}$ to estimate how much of the variation in one predictor can be predicted from others.

$$VIF_{i} = \frac{1}{1 - R^{2}_{i}}$$

---

# Gauss-Markov assumptions

Linear model must fulfill the following conditions.

- linear in parameters  
  $Y = \alpha + \beta x + \varepsilon$
- expected error is zero  
  $E(\varepsilon) = 0$
- homoscedasticity  
  $var(\varepsilon) = E(\varepsilon^{2})$
- no autocorrelation  
  $cov(\varepsilon_{i}, \varepsilon_{j}) = 0, i \neq j$
- independence of predictor(s) and residuals  
  $cov(x,\varepsilon) = 0$
  
If these are true, then the model is the *best linear unbiased estimator* (BLUE).

---

class: center middle inverse 

# Improving the functional form

---

# Use logs

When the distribution is highly skewed, we may wish to use (natural) logged values of variable(s). 

We can specify this simply in the model definition, e.g. `log(weight2) ~ weight1`

---

# Log-linear model

.pull-left[
```{r}
wtModLogLin <- lm(log(weight2) ~ weight1, calves)
summary(wtModLogLin)
```
]

.pull-right[
When 1st weight increases by 1 kg, 2nd weight increases by `r round(wtModLogLin$coef[[2]], 3)  * 100` percent.
]

---

```{r, fig.width = 16}
par(mfrow = c(1,2))
plot(wtMod, 1)
plot(wtModLogLin, 1)
```

---

# Log-log model

.pull-left[
```{r}
wtModLogLog <- lm(log(weight2) ~ log(weight1), calves)
summary(wtModLogLog)
```
]

.pull-right[
When 1st weight increases by 1 percent, 2nd weight increases by `r round(wtModLogLog$coef[[2]], 3)` percent.
]

---

# Use polynomials

Response $Y$ can be modelled as some degree polynomial in predictor $X$ . This means adding the predictor as squared, cubic, quadratic or higher term.

In the example the formula of square term added would be `weight2 ~ weight1 + I(weight1^2)`. We simply add variable `weight1` as squared and make R treat it *as is* by enclosing it in `I()`.

What this means: $Y = \alpha + \beta_{1} x  + \beta_{2} x^2+ \varepsilon$

---

.pull-left[
```{r}
wtModSq <- lm(weight2 ~ weight1 + I(weight1^2), calves)
summary(wtModSq)
```
]

.pull-right[
Did the R-squared improve?
```{r}
summary(wtMod)$r.squared
summary(wtModSq)$r.squared
```
We can't really interpret the coefficients now.
]

---

We can use `anova()` to test wether or not the squared term improves model.

```{r}
anova(wtMod, wtModSq)
```

Here the p-value of `r round(anova(wtMod, wtModSq)$P[[2]], 3)` indicates that model with squared term is not significantly different (better).

---

class: center middle inverse

# Multivariate models

---

# Let's add a predictor

Does diet also have an effect on 2nd weight? 

```{r, echo = F}
plot(weight2 ~ weight1, calves, xlab = "1st weight", ylab = "2nd weight", 
     col = rainbow(3)[calves$diet], pch = 20, cex = 2)
legend('topleft', legend = levels(calves$diet), pch = 20, pt.cex = 2, col = rainbow(3))
```

---

We can add multiple predictors to model and these can even be factors. Because all factor levels are evaluated against the first level, we may want to change it.

```{r}
levels(calves$diet)
calves$diet <- factor(calves$diet, levels = c('Low', 'Medium', 'High'))
```

---

.pull-left[
```{r}
wtMod <- lm(weight2 ~ weight1 + diet, calves)
summary(wtMod)
```
]

.pull-right[
The coefficient for 1st weight slightly increased. 

The 2nd weight if 1st weight is 0 and diet is low is `r round(wtMod$coef[[1]], 3)`.

All the values for variable `diet` are compared to base level, i.e. *Low*. 

Medium diet actually decreases 2nd weight by  `r round(wtMod$coef[[3]], 3)` compared to *low* diet. 

*High* diet is no different from *Low* as the coefficient is not statistically significant.
]

---

# Factors in OLS

.pull-left[
```{r}
summary(lm(weight2 ~ diet, calves))
```
]

.pull-right[
```{r}
aggregate(weight2 ~ diet, calves, mean)
```

Anova is equivalent to OLS with one or more factor variables.

``` {r}
summary(aov(weight2 ~ diet, calves))
```
]

---

# Why use multiple predictors?

.pull-left[
```{r}
lm(weight2 ~ weight1, calves)$coef
```

```{r}
lm(weight2 ~ weight1 + diet, calves)$coef
```
]

--

.pull-right[
Treating only diet as a predictor would have misled us.
```{r}
lm(weight2 ~ diet, calves)$coef
```

```{r}
aggregate(cbind(weight1, weight2) ~ diet, calves, mean)
```
]

---

class: inverse