---
title: Task. Data visualization
subtitle: Data and statistics with R <br> Research Methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
output:
  html_document:
    theme: yeti
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = T, warning = T, error = T)
```

# Load and understand data

The data for the exercises is on the address below. If you open it in web browser, you should see that this data is formatted as *tsv* (tab-separated value). Thus, you need to use `read.table()` command and set `sep = '\t'` to have the function treat tabs as separators. Unlike `read.csv()`, `read.table()` does not treat the first line as a header by default, so you also need to specify this with `head = T`.

https://raw.githubusercontent.com/kwstat/agridat/master/data/lehner.soybeanmold.txt

```{r, echo = F}
soybm <- agridat::lehner.soybeanmold
```

```{r, eval = F}
soybm <- read.table('https://raw.githubusercontent.com/kwstat/agridat/master/data/lehner.soybeanmold.txt', 
                    sep = '\t', head = T)
```

Once you have managed to store the data as a `data.frame` object, take a look at it with `str()` command.

``` {r}
str(soybm)
```

This is data on yield, white mold, and sclerotia fungus for soybeans in Brazil. The variables are described as follows.

- `study`, study number
- `year`, year of harvest
- `loc`, location name
- `elev`, elevation
- `region`, region
- `trt`, treatment number
- `yield`, crop yield, kg/ha
- `mold`, white mold incidence, percent
- `sclerotia`, weight of sclerotia g/ha

Looking at the raw data may give you a better understanding of it. Use `head()` to see the first 10 rows.

``` {r}
head(soybm, 10)
```

# Take a look at the distribution of mold with histogram

Suppose we are interested in the values of mold. Use `hist()` function to plot a histogram of `mold`.

```{r}
hist(soybm$mold)
```

Now plot heights of the bars as probability densities by setting `freq = F`. Also, add some more breaks with `breaks = 30`. Then add density lines with `lines(density())` and try to bring out the line by adding some color with `col = `. See the output of `colors()` to see the color names recognized by R.

```{r}
hist(soybm$mold, breaks = 30, freq = F)
lines(density(soybm$mold), col = 'tomato')
```

# Plot the relationship between mold and yield

If we were interested whether higher yields cause mold or not, we could plot these two variables against each other. Do this with `plot()` function and create a scatterplot where `mold` is on y-axis and `yield` on x-axis.

```{r}
plot(mold ~ yield, soybm)
```

Now also add a colored vertical line that indicates the mean value of yield. Use `abline(v = mean())` and add `col = ` as an argument to `abline()`.

```{r}
plot(mold ~ yield, soybm)
abline(v = mean(soybm$yield), col = 'cornflowerblue')
```

# See the change over years

Now see how the mean yield has changed. To get means for each year, pass the formula `mold ~ year` to `aggregate()` and use function `mean()` for aggregation. Store the result in a new data frame. Then reorder it to make sure it is ordered by `year`. 

```{r}
aggr <- aggregate(mold ~ year, soybm, mean)
aggr <- aggr[order(aggr$year), ]
aggr
```

Now you can use this data frame to plot `year` against `mold` as in the last plot. For time series, it is better to have a line, so set `type = 'l'`.

```{r}
plot(mold ~ year, aggr, type = 'l')
```

We can see that there has been a substatial decrease. Because the y-scale does not start from 0, this is misleading. Fix it by setting the limits of y-scale to 0 and maximum of `mold` with argument `ylim = `.

```{r}
plot(mold ~ year, aggr, type = 'l', ylim = c(0, max(aggr$mold)))
```

# Compare mean values between locations

Suppose we are now interested in how different are the mean mold incidences among locations. First you need to aggregate data to get mean vaues of `mold` by `loc`. Store the results in a data frame and then reorder it by `mold`.

```{r}
aggr <- aggregate(mold ~ loc, soybm, mean)
aggr <- aggr[order(aggr$mold), ]
aggr
```

Next, create a barplot of the aggregated values of `mold` and use `names.arg = ` to add location names. Also, rotate the plot with `horiz = T` and make labels horizontal with `las = 1`.

``` {r}
barplot(aggr$mold, names.arg = aggr$loc, horiz = T, las = 1)
```

Since the location names are quite long, they are cropped. To fix this, widen the left margin to 10 lines and keep the rest of the margins at 2 lines by adjusting plot parameters with `par(mar = c(2,10,2,2))`. These numbers control the margins in following sequence: bottom, left, top, right.

``` {r}
par(mar = c(2,10,2,2))
barplot(aggr$mold, names.arg = aggr$loc, horiz = T, las = 1)
```

# Compare yields in different locations

Now, let's see how yield compares between locations. Aggregate the data as in previous section, except this time aggregate the values of `yield` instead of `mold`. 

```{r}
aggr <- aggregate(yield ~ loc, soybm, mean)
aggr <- aggr[order(aggr$yield), ]
aggr
```

This time use `dotchart()` to plot mean values of `yield` by `loc`.

``` {r}
dotchart(aggr$yield, aggr$loc)
```

# Get a better comparison by plotting quantiles

In previous sections we only looked at the difference in mean values. However, that is only part of the picture. A better comparison can be achieved by plotting quantile ranges of different locations side by side. Do this with `boxplot()` and use the formula `mold ~ loc`. Also, add `horizontal = T` and `las = 1` to rotate the plot.

```{r, echo = -1}
par(mar = c(2,10,2,2))
boxplot(mold ~ loc, soybm, horizontal = T, las = 1)
```

# Create a publication ready plot

Suppose our last plot conveys some information we would like to publish. Thus, we need to add some elements to have a self-explanatory plot. Add the following elements:

- title, `main = `
- subtitle, `sub = `
- label for x-axis, `xlab = `
- label for y-axis, `ylab = `

```{r, echo = -1}
par(mar = c(2,10,2,2))
boxplot(mold ~ loc, soybm, horizontal = T, las = 1, 
        main = "White mold for soybeans in locations in Brazil", 
        sub = "Lehner, M. S., Pethybridge, S. J., Meyer, M. C., & Del Ponte, E. M. (2016).",
        xlab = "White mold incidence, %", 
        ylab = "")
```

Because we decreased the bottom margin, the x-axis label and subtitle are not visible. Change the bottom margin to 6 with `par(mar = c(6,10,2,2))`. Then draw the plot again.

```{r}
par(mar = c(6,10,2,2))
boxplot(mold ~ loc, soybm, horizontal = T, las = 1, 
        main = "White mold for soybeans in locations in Brazil", 
        sub = "Lehner, M. S., Pethybridge, S. J., Meyer, M. C., & Del Ponte, E. M. (2016).",
        xlab = "White mold incidence, %", 
        ylab = "")
```
