---
title: Task. Basics in R.
subtitle: Data and statistics with R <br> Research methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    highlight: default
    theme: yeti
editor_options: 
  chunk_output_type: console
---

```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = T, eval = T, echo = T, message = T, warning = T, error = T)
# Libraries
library('magrittr')
# Functions
headtail <- function(x) if (is.data.frame(x)) rbind(head(x, 3), tail(x, 3)) else c(head(x, 3), tail(x, 3))
```

The following outlines some basic functionality of R. These methods help you to understand what's in a data set, prepare data for further processing and get some simple summary statistics.

# Load data a data set into R workspace

# Have an overview of the data set

# Calculate statistics of variables

```{r}
mean(corn$flower)
```

`na.rm = TRUE`

```{r}
mean(corn$flower, na.rm = TRUE)
```

```{r}
mean(corn$yield)
```

```{r}
head(corn[corn$loc == 'Knoxville,TN', ])
```

```{r}
mean(corn[corn$loc == 'Knoxville,TN', 'yield'])
```

Now try...

- `mean()'
- `median()'
- `min()'
- `max()'
- `sum()'


# Aggregate data

Formula

```{r}
aggregate(formula = , data = , FUN = )
```

# See the distribution of values



```{r}
summary(corn$yield)
hist(corn$yield)
```

# Plot distribution of a variable by group

```{r}
boxplot(, corn)
```

# Tabulate to get counts

A very useful function in R is `table()`. Simply put, you give it a character, factor or logical variable and it returns how many times each value in that value occurs. Try to find out how many times each location (`loc`) occurs in the data set by passing the variable to `table()`. 

```{r}
table(corn$loc)
```

You can see that each location occurs 60 times, i.e. is represented by 60 rows in the data frame. 

A great feature of this function is that you can give it more than one variable. In such case, it will return a cross table. Cross tabulate genotype (`gen`) and location (`loc`) to see which genotypes are available in each location.

``` {r}
table(corn$gen, corn$loc) %>% as.data.frame.matrix %>% head
```

```{r}
is.na()
```

```{r}
table(is.na())
```

# Save data

```{r}
write.csv()
```
