---
title: Final test
subtitle: Data and statistics with R <br> Research methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    highlight: default
    theme: yeti
editor_options:
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = F, eval = F, echo = F, message = T, warning = T, error = T)
# Create data
lupin <- agridat::verbyla.lupin
lupin <- lupin[!(lupin$loc %in% c('Badgingerra', 'EastChapman', 'SouthCarrabin')), ]
lupin <- droplevels(lupin)
write.csv(lupin, '/home/jrl/rcrs/data/verbyla_lupin.csv', row.names = F)
```

For actions type respective command(s) in R. For questions provide an answer as a comment.

# Read and understand

## Read and store data

- Read data on the following *url*.
  https://raw.githubusercontent.com/lillemets/datastatsr/master/data/verbyla_lupin.csv
  
- Store this data as data frame.

You will need to use the following variables:

- `gen`, genotype, 9 varieties
- `site`, site, 11 levels
- `rep`, rep, 2-3 levels
- `rate`, seeding rate in plants/m^2
- `row`, row
- `col`, column
- `serp`, factor of 4 levels for serpentine seeding effect
- `linrow`, centered row position as a numeric variate (row-8.5)/10
- `lincol`, centered column position as a numeric variate (col-3.5)
- `linrate`, linear effect of seedrate, scaled (seedrate-41.92958)/10
- `yield`, yield in tons/hectare
- `year`, year, 1991-1992
- `loc`, location

## Understand data

- See the structure of data.

```{r}
str(lupin)
```

- See the first rows of data.

```{r}
head(lupin)
```

- See summary statistics for all variables.

```{r}
summary(lupin)
```

- How many observations are in the data?

- How many variables are of type Factor?

- How many missing values are there for variable `yield`?

# Tidy data

## Remove some observations

- Remove observations where the value for `loc` is *MtBarker*.

```{r}
lupin <- lupin[lupin$loc != 'MtBarker', ]
```

- See how many missing values are there for `yield`.

```{r}
table(is.na(lupin$yield))
```

- Remove observations where `yield` is missing.

```{r}
lupin <- lupin[!is.na(lupin$yield), ]
```

## Add variables

- Add variable called `place` that has value *"Newdegate"* when `loc` is *Newdegate* and when it is something else then *"Others"*.

```{r}
lupin$place <- ifelse(lupin$loc == 'Newdegate', 'Newdegate', 'Others')
```

## Replace a variable

- `yield` is represented in t/ha. Replace the values of `yield` so that it would be kg/ha. You need to multiply values of `yield` by 1000.

```{r}
lupin$yield <- lupin$yield * 1000
```

## Tabulate

- Count the observations of `loc` in both groups of `place`.

```{r}
table(lupin$loc, lupin$place)
```

# Descriptive statistics

## Describe the distribution

- Find the mean, median, minimum and maximum values of `yield`.

```{r}
summary(lupin$yield)
```

- Plot the distribution of `yield`.

```{r}
hist(lupin$yield)
```

## Summarize by groups

- Get mean values of `yield` by `place`.

```{r}
aggregate(yield ~ place, lupin, mean)
```

- Find mean values of `yield` for each `loc`.

```{r}
aggregate(yield ~ loc, lupin, mean)
```

- Draw boxplots of the values of `yield` of each `loc`.

```{r}
boxplot(yield ~ loc, lupin)
```

# Test differences in means

- Test if mean yield is different for *Newdegate* and *Others* in variable `place`.

```{r}
t.test(yield ~ place, lupin)
```

- Is there a difference?

- Test if mean yields differ by `loc`.

```{r}
summary(aov(yield ~ loc, lupin))
```

- Is any of the environments in `loc` different from others in terms of yield?

# Correlations

## Correlation coefficient

- Plot the relationship between `yield` and `rate`.

```{r}
plot(yield ~ rate, lupin)
```

- Find the Pearson's correlation coefficient between `yield` and `rate`.

```{r}
cor(lupin$yield, lupin$rate, use = 'pairwise.complete.obs')
```

- How would you describe this relationship?

# Regression modelling

## Estimate an OLS regression model

- Estimate an OLS regression model where `yield` depends on `height`.

```{r}
ylMod <- lm(yield ~ height, lupin)
summary(ylMod)
```

- What is the theoretical yield when height is 0?

- How many units yield increases when height increases by 1 unit?

- How much of the variance in yield is explained by the model?

## See how well the model fits

- Plot the relationship between `yield` and `height` and add the regression line of the model.

```{r}
plot(yield ~ height, lupin)
abline(ylMod, col = 'tomato')
```

- Deciding by the plot, does the model fit data well? 

- Plot residuals against fitted values.

```{r}
plot(ylMod, 1)
```

- Is the variance of the residuals constant?

- Plot QQ-plot of standardized residuals

```{r}
plot(ylMod, 2)
```

- Are residuals normally distributed?

## Estimate a multivariate OLS regression model

- Estimate another model where you also add `hddate` and `WA` as predictors.

```{r}
ylMod2 <- lm(yield ~ height + hddate + wa, lupin)
summary(ylMod2)
```

- How do you interpret the intercept now?

- How do you interpret the coefficient of `wa`?

- Test if models are different.

```{r}
anova(ylMod, ylMod2)
```

- Is the second model different from first? How?