---
title: Revision for final test
subtitle: Data and statistics with R <br> Research methods <br><br>
author: Jüri Lillemets
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    highlight: default
    theme: yeti
editor_options:
  chunk_output_type: console
---
  
```{r setup, include = F}
# Settings
knitr::opts_chunk$set(include = F, eval = F, echo = F, message = T, warning = T, error = T)
# Create data
barley <- agridat::steptoe.morex.pheno
write.csv(barley, '/home/jrl/rcrs/data/steptoe_morex_pheno.csv', row.names = F)
```

For actions type respective command(s) in R. For questions provide an answer as a comment.

# Read and understand

## Read and store data

- Read data on the following *url*.
  https://raw.githubusercontent.com/lillemets/datastatsr/master/data/steptoe_morex_pheno.csv
  
- Store this data as data frame.

The data has the following variables:

- `gen`, genotype factor...
- `env`, environment, 16 levels
- `amylase`, alpha amylase (20 Deg Units)
- `diapow`, diastatic power (degree units) 
- `hddate`, heading date (julian days)
- `lodging`, lodging (percent)
- `malt`, malt extract (percent)
- `height`, plant height (centimeters)
- `protein`, grain protein (percent)
- `yield`, grain yield (Mt/Ha)

## Understand data

- See the structure of data.

- See the first rows of data.

- See summary statistics for all variables.

- How many observations are in the data?

- How many variables are of type Factor?

- How many missing values are there for variable `logding`?

```{r}
str(barley)
head(barley)
summary(barley)
```

# Tidy data

## Remove some observations

- Remove observations where the value for `env` is *ON92*.

```{r}
barley <- barley[barley$env != 'ON92', ]
```

- See how many missing values are there for `height`.

```{r}
table(is.na(barley$height))
```

- Remove observations where `height` is missing.

```{r}
barley <- barley[!is.na(barley$height), ]
```

## Add variables

- Add variable called `wa` that has value *"WA"* when `env` is *WA91* or *WA92* and when it is something else then *"Others"*.

```{r}
barley$wa <- ifelse(barley$env == 'WA91' | barley$env == 'WA92', 'WA', 'Others')
```

- Add variable called `high.prot` using condition where `protein` is greater than mean of `protein`.

``` {r}
barley$high.prot <- barley$protein > mean(barley$protein, na.rm = T)
```

## Replace a variable

- `yield` is represented in tons/ha. Replace the values of `yield` so that it would be kg/ha. You need to multiply values of `yield` by 1000.

```{r}
barley$yield <- barley$yield * 1000
```

## Tabulate

- Count the observations of `env` in both groups of `wa`.

```{r}
table(barley$env, barley$wa)
```

- See how many values of `protein` are missing for each `env`.

``` {r}
table(barley$env, is.na(barley$protein))
```

# Descriptive statistics

## Describe the distribution

- Find the mean, median, minimum and maximum values of `yield`.

```{r}
summary(barley$yield)
```

- Plot the distribution of `protein`.

```{r}
hist(barley$protein)
```

## Summarize by groups

- Get mean values of `yield` by `wa`.

```{r}
aggregate(yield ~ wa, barley, mean)
```

- Find mean values of `protein` for each `env`.

```{r}
aggregate(protein ~ env, barley, mean)
```

- Draw boxplots of the values of `protein` of each `env`.

```{r}
boxplot(protein ~ env, barley)
```

# Test differences in means

- Test if mean yield is different for *WA* and *Other* locations in variable `wa`.
- Is there a difference?

```{r}
t.test(yield ~ wa, barley)
```

- Test if mean yield of *WA* is higher than mean yield of other locations.
- Is it higher?

```{r}
unique(barley$wa)
t.test(yield ~ wa, barley, alt = 'less')
```

- Test if mean yields differ by `env`.
- Is any of the environments in `env` different from others in terms of yield?

```{r}
summary(aov(protein ~ env, barley))
```

# Correlations

## Correlation coefficient

- Plot the relationship between `yield` and `height`.

```{r}
plot(yield ~ height, barley)
```

- Find the Pearson's correlation coefficient between `yield` and `height`.

```{r}
cor(barley$yield, barley$height, use = 'pairwise.complete.obs')
```

- How would you describe this relationship?

## Heatmap

- Create a heatmap of the values of Pearson's correlation coefficient between all numeric variables.

```{r}
heatmap(cor(barley[, 3:10], use = 'pairwise.complete.obs'))
```

- Which variables are most strongly correlated with `height`?

# Regression modelling

## Estimate an OLS regression model

- Estimate an OLS regression model where `yield` depends on `height`.

```{r}
ylMod <- lm(yield ~ height, barley)
summary(ylMod)
```

- What is the theoretical yield when height is 0?

- How many units yield increases when height increases by 1 unit?

- How much of the variance in yield is explained by the model?

## See how well the model fits

- Plot the relationship between `yield` and `height` and add the regression line of the model.

```{r}
plot(yield ~ height, barley)
abline(ylMod, col = 'tomato')
```

- Deciding by the plot, does the model fit data well? 

- Plot residuals against fitted values.

```{r}
plot(ylMod, 1)
```

- Is the variance of the residuals constant?

- Plot QQ-plot of standardized residuals

```{r}
plot(ylMod, 2)
```

- Are residuals normally distributed?

## Estimate a multivariate OLS regression model

- Estimate another model where you also add `hddate` and `WA` as predictors.

```{r}
ylMod2 <- lm(yield ~ height + hddate + wa, barley)
summary(ylMod2)
```

- How do you interpret the intercept now?

- How do you interpret the coefficient of `wa`?

- Test if models are different.

```{r}
anova(ylMod, ylMod2)
```

- Is the second model different from first? How?

## Estimate a logistic regression model

- Estimate a logistic regression model where `high.prot` is the response and `height`, `hddate` and `wa ` are predictors.

```{r}
summary(glm(high.prot ~ height + hddate + wa, barley, family = 'binomial'))
```

- How do you interpret the coefficients for `height`, `hddate`, and `wa`?